Acceleration of Derivative Calculations with Application to Radial Basis Function ­ Finite-Differences on the Intel MIC Architecture

### 2014-06-12 slides/2014_ics/slides-ics.tex
slides

@@

\section{Introduction}

\subsection{RBF-FD}

\begin{frame}[fragile]
  \frametitle{Derivative as linear algebra operators}

@@

\begin{frame}
  \frametitle{RBF-FD methods}

  \todo{Add definition of RBF}

  \begin{columns}
    \column{.45\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/RBFStencil_n32_a}

    An example of an RBF-FD stencil of size $n=32$ on a sphere of
    $N=1024$ nodes to approximate any derivative operator at the point
    in the square.

    \column{.55\linewidth}

    \begin{block}{Shallow waters equations of a rotating fluid}
      \tiny
      \begin{eqnarray}
        \dfrac{\partial u}{\partial t} &=& - \left( u\dfrac{\partial u}{\partial x} + v\dfrac{\partial u}{\partial y} + w\dfrac{\partial u}{\partial z} + f(yw - zv) + g\dfrac{\partial h}{\partial x} \right) 
        \nonumber \\
        \dfrac{\partial v}{\partial t} &=& - \left( u\dfrac{\partial v}{\partial x} + v\dfrac{\partial v}{\partial y} + w\dfrac{\partial v}{\partial z} + f(zu - xw) + g\dfrac{\partial h}{\partial x} \right)
        \nonumber\\
        \dfrac{\partial w}{\partial t} &=& - \left( u\dfrac{\partial w}{\partial x} + v\dfrac{\partial w}{\partial y} + w\dfrac{\partial w}{\partial z} + f(xv - yu) + g\dfrac{\partial h}{\partial x} \right)
        \nonumber\\
        \frac{\partial h}{\partial t}& =&-\left(\dfrac{\partial (uh)}{\partial x} + \dfrac{\partial (vh)}{\partial y} + h\dfrac{\partial (wh)}{\partial z}\right) , \label{height} \nonumber
      \end{eqnarray}
    \end{block}

    How to compute one simple derivative: sparse matrix
    multiplication. Bound by (roughly) 25GFlop/s. Can we do better?
  \end{columns}
\end{frame}

\subsection{MIC}

% \begin{frame}
%   \frametitle{What is Intel MIC}
%   Intel Many integrated Core (MIC) is Intel's response to GPUs
%   becoming popular in High Performance Computing.

%   \begin{block}{What GPUs do well?}
%     \begin{itemize}
%       \item Get a lot of GFlops by using hundreds of cores
%       \item Each core has large SIMD-like abilities
%       \item Hide memory latency by using fast cycle context switch
%     \end{itemize}
%   \end{block}

%   \begin{block}{What GPUs do not do well?}
%     \begin{itemize}
%       \item Alien to program
%       \item Poor support for legacy applications
%       \item Inter thread communications
%       \item Branching
%     \end{itemize}
%   \end{block}

%   Goal of \MIC : do all of it well!
% \end{frame}

\begin{frame}
  \frametitle{Intel MIC: Overall Architecture}
  \begin{columns}
    \column{18em}    
    \includegraphics[width=18em]{slides-figures/MIC-overview.pdf}
    \column{17em}

    \begin{itemize}
      \item 8 memory controllers
        \begin{itemize}
          \item GDDR5
          \item 2 channels (32-bit)
          \item 5.5GT/s
          \item 352GB/s aggregated peak
            \begin{itemize}
              \item about same as a K40 
              \item but you typically get half
            \end{itemize}
        \end{itemize}
      \item 50+ cores
        \begin{itemize}
          \item 32KB of L1 cache
          \item 512KB of L2 cache
          \item LRU
          \item 8-way associative
        \end{itemize}
      \item 1 PCI-e controller
        \begin{itemize}
          \item to the host (2GB/s guaranteed to memory)
        \end{itemize}
    \end{itemize}
  \end{columns}

@@

\begin{frame}
  \frametitle{Core architecture}
  \begin{columns}
    \column{20em}
    \includegraphics[width=20em,height=20em]{slides-figures/MIC-core.png}%from intelsdg  
    \column{12em}
    \begin{itemize}
      \small
    \item Clocked at 1Ghz
    \item 64-bit
    \item 4 hardware threads
      \begin{itemize}
        \footnotesize
        \item no context switching
        \item no 2 instructions for the same thread
      \end{itemize}
    \item A vectorial unit
      \begin{itemize}
        \footnotesize
        \item 512-bit registers
        \item 16 SP floats
        \item 8 DP floats
        \item support FMA
      \end{itemize}
    \item Two instruction pipes:
      \begin{itemize}
        \footnotesize
        \item 2 ALU ops
        \item ALU + MEM ops
        \item ALU + VFP ops
        \item VFP + MEM ops
      \end{itemize}
    \item In-order execution
    \end{itemize}
  \end{columns}
\end{frame}

% \begin{frame}
%   \frametitle{Vectorial Unit (SIMD) : The classics}

%   \begin{columns}
%     \column{15em}

%     \includegraphics[width=15em]{slides-figures/vect-basic.png}%from intelsdg

%     \column{15em}

%     {\small 32 512-bits registers (16x32, 8x64)}

%     \begin{block}{Binary operators (1 cycle throughput)}
%       \small mul, add, sub, and, xor, or, fma
%     \end{block}

%     \begin{block}{Short Vector Math Library}
%       sqrt, rsqrt, log, exp, div
%     \end{block}

%   \end{columns}
% \end{frame}

\begin{frame}
  \frametitle{When Should I Use \MIC ?}

  \begin{block}{Key points of SE10P}
    \begin{itemize}
      \item Large memory bandwidth (peak: 220GB/s)
      \item 61 cores with mandatory use of hardware threading
      \item 512-bit wide SIMD registers: 
        \begin{itemize} 
          \item FMA: up to 2x16 SP Flop/c (2x8 DP Flop/c)
          \item otherwise: up to 16 SP Flop/c (8 DP Flop/c)
        \end{itemize}
      \item On a 61-core configuration at 1.05Ghz:
        \begin{itemize} 
        \item FMA: 2x16x61x1.05Ghz = 2.048 TFlop/s SP (1.024TFlop/s DP)
        \item otherwise: 16x61x1.05Ghz = 1.024 GFlop/s SP (512GFlop/s DP)
        \end{itemize}
    \end{itemize}
  \end{block}

  Sounds like the perfect system for Sparse Matrix operations!
\end{frame}

% \begin{frame}
%   \frametitle{Have to use vector operations to get high bandwidth}

%   \begin{columns}
%     \column{.49\linewidth}

%     \begin{center}
%       \includegraphics[width=\linewidth,page=1]{slides-figures/PPAM-data/bandwidth.pdf}

%       Read
%     \end{center}

%     \column{.49\linewidth}

%     \begin{center}
%       \includegraphics[width=\linewidth,page=2]{slides-figures/PPAM-data/bandwidth.pdf}

%       Write
%     \end{center}

%   \end{columns}

%   \begin{itemize}
%   \item Using the appropriate vectorial instructions gives significant
%     improvements.
%   \item Read Peak: 183GB/s.
%   \item Write Peak: 160GB/s.
%   \end{itemize}

%   \tiny Source: PPAM 2013
% \end{frame}

\begin{frame}
  \frametitle{Compute all the simple derivative at once}

%  \begin{columns}
%    \column{.5\linewidth}
    \begin{block}{Idea}
      \scriptsize
      \begin{equation}
        \left( \begin{array}{cccc}
            \underline{u}_x     & \underline{v}_x     & \underline{w}_x     & \underline{h}_x \\
            \underline{u}_y     & \underline{v}_y     & \underline{w}_y     & \underline{h}_y \\
            \underline{u}_z     & \underline{v}_z     & \underline{w}_z     & \underline{h}_z \\
            \underline{u}_{hyp} & \underline{v}_{hyp} & \underline{w}_{hyp} & \underline{h}_{hyp}
          \end{array} \right)
        = \left(
          \begin{array}{c}
            D_{x}\\ D_{y}\\ D_{z}\\ D_{hyp}
          \end{array}\right)
        \times \left(\underline{u} \,\, \underline{v} \,\, \underline{w} \,\, \underline{h} \,\,\right) 
        \vspace{-1ex}
      \end{equation}
    \end{block}  
%    \column{.5\linewidth}

    \begin{block}{Why would it work}
      All the derivative matrices have the same sparsity pattern. In
      other words, the non zeros are at the same positions in the
      matrix, only the values of the non zero change.
    \end{block}

%  \end{columns}
\end{frame}

@@

  \begin{columns}
    \column{.49\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_best-crop.pdf}

    \begin{center}
      Best case\\~
    \end{center}

    \column{.49\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_worst-crop.pdf}
    \begin{center}
      Worst case\\~
    \end{center}

    % \column{.33\linewidth}
    % \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_no_cache-crop.pdf}
    % \begin{center}
    %   Worst case \\no cacheline
    % \end{center}
  \end{columns}

  \tiny Also, taking cacheline into account makes a large difference (see paper for details).
\end{frame}

@@

\section{Implementation (4 vectors, 4 matrices)}

\begin{frame}
  \frametitle{Observation on the 4 vectors, 4 matrices case}

  \begin{itemize}
  \item We aim at over 200Gflop/s in single precision. (\~{}  10\% of peak.)
  \item One instruction in ten must be a Fused Multiply Add.
  \item The kernel must be fairly instruction efficient
  \end{itemize}
\end{frame}

@@

    \begin{block}{RBF-FD}
      \centering
      \includegraphics[width=.6\linewidth]{slides-figures/ICS-figures/kd-tree-3d-norcm-crop.png}
    \end{block}

@@

>>>>

\section{Radial Basis Function - Finite Difference}

@@

\begin{frame}[fragile]
  \frametitle{Derivative as (sparse) linear algebra operators}

@@

\begin{frame}[fragile]
  \frametitle{Radial Basis Function}
  \begin{columns}
    \column{.45\linewidth}

    \begin{block}{RBF}
      \includegraphics[width=\linewidth]{slides-figures/rbf.pdf}

      \begin{itemize}
      \item Radial symmetry.
      \item RBFs approximate a function $f$ sampled at N distinct location $$f(\vec{r}) = \sum_i w_i \phi(||\vec{r}-\vec{r_i}||)$$ 
      \end{itemize}
    \end{block}

    \column{.45\linewidth}
    \begin{block}{The RBF-FD method}
      \includegraphics[width=.8\linewidth]{slides-figures/ICS-figures/RBFStencil_n32_a}

      Derivative operators of an RBF are approximated using the $n=32$
      closest sampling points on a sphere of $N=1024$ points.

      The DM has exactly $n$ non zero per row.
    \end{block}

\begin{frame}
  \frametitle{Computing derivatives. One at a time?}

  \begin{block}{Shallow waters equations}

    \begin{eqnarray}
      \dfrac{\partial u}{\partial t} &=& - \left( u\dfrac{\partial u}{\partial x} + v\dfrac{\partial u}{\partial y} + w\dfrac{\partial u}{\partial z} + f(yw - zv) + g\dfrac{\partial h}{\partial x} \right) 
      \nonumber \\
      \dfrac{\partial v}{\partial t} &=& - \left( u\dfrac{\partial v}{\partial x} + v\dfrac{\partial v}{\partial y} + w\dfrac{\partial v}{\partial z} + f(zu - xw) + g\dfrac{\partial h}{\partial y} \right)
      \nonumber\\
      \dfrac{\partial w}{\partial t} &=& - \left( u\dfrac{\partial w}{\partial x} + v\dfrac{\partial w}{\partial y} + w\dfrac{\partial w}{\partial z} + f(xv - yu) + g\dfrac{\partial h}{\partial z} \right) 
      \nonumber\\
      \frac{\partial h}{\partial t} &=& -\left(\dfrac{\partial (uh)}{\partial x} + \dfrac{\partial (vh)}{\partial y} + h\dfrac{\partial (wh)}{\partial z}\right)  \nonumber
    \end{eqnarray}
    $u,v,w$ velocity. $h$ geopotential height.
  \end{block}

  One at a time? $\dfrac{\partial u}{\partial x}$, then $\dfrac{\partial v}{\partial y}$

  \only<2>{\vspace{-19em}\hspace{-1em}\begin{minipage}{.95\linewidth}
      \colorbox{white}{
        \includegraphics[width=\linewidth]{slides-figures/PPAM-figures/all_compare_spmv.pdf}
        \tiny PPAM 2013
      }
    \end{minipage}}

@@

\begin{frame}
  \frametitle{Compute all the simple derivatives at once}

\only<1-2>{  \begin{block}{Shallow waters equations}
    \begin{eqnarray}
      \dfrac{\partial u}{\partial t} &=& - \left( u\dfrac{ \textcolor{blue}{\partial}  u}{\textcolor{blue}{\partial x}} + v\dfrac{\textcolor{red}{\partial} u}{\textcolor{red}{\partial y}} + w\dfrac{\textcolor{green}{\partial} u}{\textcolor{green}{\partial z}} + f(yw - zv) + g\dfrac{\textcolor{blue}{\partial} h}{\textcolor{blue}{\partial x}} \right) 
      \nonumber \\
      \dfrac{\partial v}{\partial t} &=& - \left( u\dfrac{ \textcolor{blue}{\partial} v}{\textcolor{blue}{\partial x}} + v\dfrac{\textcolor{red}{\partial} v}{\textcolor{red}{\partial y}} + w\dfrac{\textcolor{green}{\partial} v}{\textcolor{green}{\partial z}} + f(zu - xw) + g\dfrac{\textcolor{red}{\partial} h}{\textcolor{red}{\partial} y} \right)
      \nonumber \\
      \dfrac{\partial w}{\partial t} &=& - \left( u\dfrac{\textcolor{blue}{\partial} w}{\textcolor{blue}{\partial x}} + v\dfrac{\textcolor{red}{\partial} w}{\textcolor{red}{\partial y}} + w\dfrac{\textcolor{green}{\partial} w}{\textcolor{green}{\partial z}} + f(xv - yu) + g\dfrac{\textcolor{green}{\partial} h}{\textcolor{green}{\partial z}} \right) 
      \nonumber\\
      \frac{\partial h}{\partial t} &=& -\left(\dfrac{\partial (uh)}{\partial x} + \dfrac{\partial (vh)}{\partial y} + h\dfrac{\partial (wh)}{\partial z}\right)  \nonumber
    \end{eqnarray}
  \end{block}}

\pause

  \begin{block}{Idea}
    \scriptsize
    \begin{equation}
      \left( \begin{array}{cccc}
          \underline{u}_x     & \underline{v}_x     & \underline{w}_x     & \underline{h}_x \\
          \underline{u}_y     & \underline{v}_y     & \underline{w}_y     & \underline{h}_y \\
          \underline{u}_z     & \underline{v}_z     & \underline{w}_z     & \underline{h}_z \\
          \underline{u}_{hyp} & \underline{v}_{hyp} & \underline{w}_{hyp} & \underline{h}_{hyp}
        \end{array} \right)
      = \left(
        \begin{array}{c}
          \textcolor{blue}{D_{x}}\\ \textcolor{red}{D_{y}}\\ \textcolor{green}{D_{z}}\\ D_{hyp}
        \end{array}\right)
      \times \left(\underline{u} \,\, \underline{v} \,\, \underline{w} \,\, \underline{h} \,\,\right) \nonumber
      \vspace{-1ex}
    \end{equation}
  \end{block}

\pause

  \begin{block}{Why would it work}
    All the derivative matrices have the same sparsity pattern.

    In other words, the non zeros are at the same positions in the
    matrix, only the values of the non zero change.

    Benefits:
    \begin{itemize}
    \item Matrix compression: 1 column ID, 4 nonzero value. (gain: 38\%)
    \item Vector reusage.
    \end{itemize}
  \end{block}
\end{frame}

@@

  \begin{columns}
\only<1>{    \column{.49\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_best-crop.pdf}

    \begin{center}
      Best case\\~
    \end{center}
}
    \column{.49\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_worst-crop.pdf}
    \begin{center}
      Worst case\\~
    \end{center}

\only<2>{
    \column{.49\linewidth}
    \includegraphics[width=\linewidth]{slides-figures/ICS-figures/flops_to_bytes_no_cache-crop.pdf}
    \begin{center}
      Worst case \\no cacheline
    \end{center}
}
  \end{columns}

%  \tiny Also, taking cacheline into account makes a large difference (see paper for details).
\end{frame}

@@

\section{Implementation on Xeon Phi (4 vectors, 4 matrices, Single Precision)}

% \begin{frame}
%   \frametitle{Intel MIC: Overall Architecture}
%   \begin{columns}
%     \column{18em}    
%     \includegraphics[width=18em]{slides-figures/MIC-overview.pdf}
%     \column{17em}

%     \begin{itemize}
%       \item 8 memory controllers
%         \begin{itemize}
%           \item GDDR5
%           \item 2 channels (32-bit)
%           \item 5.5GT/s
%           \item 352GB/s aggregated peak
%             \begin{itemize}
%               \item about same as a K40 
%               \item but you typically get half
%             \end{itemize}
%         \end{itemize}
%       \item 50+ cores
%         \begin{itemize}
%           \item 32KB of L1 cache
%           \item 512KB of L2 cache
%           \item LRU
%           \item 8-way associative
%         \end{itemize}
%       \item 1 PCI-e controller
%         \begin{itemize}
%           \item to the host (2GB/s guaranteed to memory)
%         \end{itemize}
%     \end{itemize}
%   \end{columns}

% \end{frame}

% \begin{frame}
%   \frametitle{Core architecture}
%   \begin{columns}
%     \column{20em}
%     \includegraphics[width=20em,height=20em]{slides-figures/MIC-core.png}%from intelsdg  
%     \column{12em}
%     \begin{itemize}
%       \small
%     \item Clocked at 1Ghz
%     \item 64-bit
%     \item 4 hardware threads
%       \begin{itemize}
%         \footnotesize
%         \item no context switching
%         \item no 2 instructions for the same thread
%       \end{itemize}
%     \item A vectorial unit
%       \begin{itemize}
%         \footnotesize
%         \item 512-bit registers
%         \item 16 SP floats
%         \item 8 DP floats
%         \item support FMA
%       \end{itemize}
%     \item Two instruction pipes:
%       \begin{itemize}
%         \footnotesize
%         \item 2 ALU ops
%         \item ALU + MEM ops
%         \item ALU + VFP ops
%         \item VFP + MEM ops
%       \end{itemize}
%     \item In-order execution
%     \end{itemize}
%   \end{columns}
% \end{frame}

\begin{frame}
  \frametitle{Summary of \MIC}

  \begin{block}{Key points of Xeon Phi SE10P}
    \begin{itemize}
    \item Accelerator card (soon main processor with Knights Landing) 
    \item Large memory bandwidth (peak: 220GB/s)
    \item 61 cores with mandatory use of hardware threading
    \item Mandatory use of 512-bit wide SIMD registers: 
      \begin{itemize} 
      \item FMA: up to 2x16 SP Flop/c (2x8 DP Flop/c)
      \item otherwise: up to 16 SP Flop/c (8 DP Flop/c)
      \end{itemize}
    \item On a 61-core configuration at 1.05Ghz:
      \begin{itemize} 
      \item FMA: 2x16x61x1.05Ghz = 2.048 TFlop/s SP (1.024TFlop/s DP)
      \item otherwise: 16x61x1.05Ghz = 1.024 GFlop/s SP (512GFlop/s DP)
      \end{itemize}
    \end{itemize}
  \end{block}

  Sounds like the perfect system for scientific computing!
\end{frame}

\begin{frame}
  \frametitle{Observations on the 4 vectors, 4 matrices case}

  \begin{itemize}
  \item We aim at over 200Gflop/s in single precision.
  \item Peak of Xeon Phi is 2TFlop/s.
  \item We want 10\% of peak performance.
  \item One instruction in ten must be a Fused Multiply Add.
  \item The kernel must be fairly instruction-efficient.
  \end{itemize}
\end{frame}

@@

    \begin{block}{RBF-FD}
      \centering
      \includegraphics[width=.6\linewidth]{slides-figures/ICS-figures/kd-tree-3d-rcm-crop.png}
    \end{block}

@@

\begin{frame}
  \frametitle{Thank you}

  \begin{block}{Other related works}
    \begin{itemize}
    \item Scalability of graph algorithms on Intel MIC. (MTAAP 2012)
    \item Sparse Matrix Vector Multiplication on Xeon Phi. (PPAM 2013)
    \item Vectorization of Graph Centrality on Xeon Phi. (MTAAP 2014)
    \end{itemize}
  \end{block}

  \begin{block}{More information}
    Contact : esaule@uncc.edu

    Visit: \url{http://webpages.uncc.edu/~esaule}
  \end{block}
\end{frame}
